---
title: "1. Lubridate for Dates"
author: "Dom Grisafe"
date: "2023-01-26"
output: rmarkdown::github_document
---

# Question

> "I have a data set that has the date format as: 11/4/2021  7:00:00 AM. 
>
> "I’d like to get it to just month and year: MM/YY like this."

# Answer

We'll use the [lubridate](https://cran.r-project.org/web/packages/lubridate/index.html) package.

Here’s the appropriate section in [R4DS](https://bookdown.org/roy_schumacher/r4ds/dates-and-times.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
I like to load the tidyverse master library in the beginning using `library(tidyverse)`.

The tidyverse library includes many packages often used, including libraries for lubridate and dplyr.


## Before we get started: `glimpse()` function

The `glimpse()` function allows us to search any object in R. It is in the dplyr package, an important library in the "Tidyverse".

I sometimes include the package name before a function for learning purposes to emphasize which library the function comes from.

For example, I could write `dplyr:glimpse()`

But from here forward I'll use `glimpse()`


## Simulating the starting data

We'll create a vector of dates formatted as they were in the initial question: `11/4/2021  7:00:00 AM`

I arbitrarily chose to make a vector of 20 dates for this example.

Don't worry about how we're doing this for now; it's not important for learning how to answer our question.

```{r, echo=FALSE}
# function from: https://stackoverflow.com/questions/14720983/efficiently-generate-a-random-sample-of-times-and-dates-between-two-dates
latemail <- function(N, st="1995/01/01", et="2022/12/31") {
  st <- as.POSIXct(as.Date(st))
  et <- as.POSIXct(as.Date(et))
  dt <- as.numeric(difftime(et,st,unit="sec"))
  ev <- sort(runif(N, 0, dt))
  rt <- st + ev
}
# print(latemail(10))

# format the date object how the question was asked: https://www.statology.org/r-date-format/
sample_dates <- latemail(20) %>% format("%m/%d/%y %I:%M:%S %p")
sample_dates %>% glimpse
```


## Whats the class?

If we look at our vector, we can see which "class" it has using the `class()` function.

```{r}
# show the current class before parsing
sample_dates %>% class()
```

A character class is commonly used in R. It is basically text. It is sometimes called a "string" in other programming languages. But in R we call in "character"

## POSIXct: the class of date and time

Dates are handled best in R using the "POSIXct" class, which is basically an integer with a starting point (e.g., the birth of Christ in year 0, or some other reference point I cannot recall).

Let's use the lubridate package, specifically the `parse_date_time()` function which will convert our "character" class to the "POSIXct" class.

```{r}
# convert character format into date format (POSIXct)
sample_dates_posixct <- sample_dates %>% lubridate::parse_date_time("%m/%d/%y %I:%M:%S %p")
sample_dates_posixct %>% glimpse
```

After taking a glimpse at our new vector, we see the old character vector is now in the standard "POSIXct" format, which is easy to work with.

Let's check the class specifically again using the `class()` function.

```{r}
# show the new class after parsing
sample_dates_posixct %>% class()
```

But how did we tell the `parse_date_time()` function to interpret the character dates correctly into the POSIXct class?


### Formatting date and time in R

We had to supply the formats of our character vector to the function `lubridate:parse_date_time()`

We told the lubridate function our old character vector was in the following format: `%m/%d/%y %I:%M:%S %p`

All date and time formats can be illustrated in R by searching the documentation of the `strptime()` function.

From the documentation, we can see what each piece of the above line means:

`%m` is for "Month as decimal number (01–12)"

`%d` is for "Day of the month as decimal number (01–31)"

See if you can search the documentation for `strptime()` to see what the other codes mean.


### Searching the documentation

Call the documentation for any function by typing a question mark before it, without parenthesis: `?strptime()`

This works for any function. Try it out with some of the other functions we've used above:

`?class()`

`?glimpse()`

Reading the documentation is an essential R skill. I use it often to determine how to use a specific function.


## Outputting the date as month and year

So our original goal was to convert these dates to just the month and year. 

We can use the lubridate package to create separate vectors of just the month and year.

This is done simply using the `lubridate:month()` and `lubridate:year()` functions.

```{r}
sample_dates_months <- sample_dates_posixct %>% lubridate::month()
sample_dates_months %>% glimpse

sample_dates_years <- sample_dates_posixct %>% lubridate::year()
sample_dates_years %>% glimpse
```

But we wanted a variable that was month and year

```{r}
sample_dates_mos_year <- paste0(sample_dates_months, sep = "/", sample_dates_years)
sample_dates_mos_year %>% glimpse
```

Check out the documentation for the function `paste0()`

See if you can figure out how it works.


## Tibble: putting it all together

Now that we have all these vectors, lets wrap them all up into a tibble, a commonly used object in R. A tibble used to be a data frame, but this is the upgraded version.

```{r}
tibble(
  question = sample_dates,
  answer = sample_dates_mos_year
)
```

